---
title: "final_report_Daniel"
author: "DSW"
date: "2025-04-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Data Exploration

To better understand the dynamics of electricity consumption and generation in the ERCOT region, we wanted to visualize our time series data (using the autoplot() function). These visuals gave us insights into long-term trends and seasonal patterns in ERCOT load, temperature, and the evolving fuel mix. By looking at these trends, we could spot times of high demand and see how temperature changes affect energy use. We also noticed shifts in the fuel mix, showing how the region is moving toward more sustainable energy sources.

## ERCOT Load

The ERCOT load has shown a steady increase from 2017 to 2025. This rise reflects growing residential and industrial electricity consumption, which could be attributed to population growth, economic development, and infrastructure expansion across Texas. Seasonal fluctuations are also apparent, with noticeable peaks during periods of extreme temperatures, particularly in the summer and winter months.

```{r TS ERCOT Load Plot}
autoplot(ts_load, series = "ERCOT Load") +
  ggtitle("ERCOT Load") +
  xlab("Time") +
  ylab("Load (MW)") +
  theme_minimal() +
  guides(colour = guide_legend(title = "Series"))
```

The ACF of the ERCOT load series shows a slow decay, suggesting non-stationarity and a strong trend over time. The irregular PACF, with spikes at various lags, points to the seasonal components in the data, and it gives some insight into how seasonality can be incroporated into an auto-regressive model. The results of the ACF and PACF plots align with our visualizations, which reveal a steady rise in load from 2017 to 2025 and clear seasonal peaks during periods of extreme temperatures. These patterns show the impact of long-term growth and seasonal demand shifts in Texas's electricity consumption.

```{r Load ACF & PACF}
# ACF and PACF for ERCOT Load
ggAcf(ts_load, lag.max = 365) + ggtitle("ACF of ERCOT Load") + theme_minimal()
ggPacf(ts_load, lag.max = 365) + ggtitle("PACF of ERCOT Load") + theme_minimal()
```

## ERCOT Temperature

As mentioned before, the temperature data was averaged daily for each zone and then combined to get an overall ERCOT-wide daily average. This change helped us match the temperature data with load and fuel mix data. As expected, temperature patterns showed strong seasonality, with summer highs and winter lows corresponding to fluctuations in load. 

```{r Temp TS}
autoplot(ts_temp) +
  ggtitle("ERCOT Area Temperature") +
  xlab("Time") +
  ylab("Temperature (°F)") +
  theme_minimal()
```

The ERCOT temperature series also exhibits a decaying ACF, indicating non-stationarity and potential long-term trends in temperature over time. The PACF of the temperature is similar to the PACF of the load, with an initial short decay over the first 10 lags followed by scattered significant spikes, suggesting both short-term autocorrelation and possible seasonal components. These results seemed to match up with our data for the ERCOT load, meaning it may have a significant role as an exogenous variable in scenario generation.


```{r Temp ACF & PACF}
# ACF and PACF for ERCOT Temperature
ggAcf(ts_temp, lag.max = 365) + ggtitle("ACF of ERCOT Temperature") + theme_minimal()
ggPacf(ts_temp, lag.max = 365) + ggtitle("PACF of ERCOT Temperature") + theme_minimal()
```

## ERCOT Fuel Mix

ERCOT’s fuel mix shows significant shifts over time. Natural gas remains the most relevant and responsive energy source, with a load shape closely mirroring the overall ERCOT demand curve. Coal and lignite usage has declined steadily, and this aligns with national efforts to reduce reliance on fossil fuels, as renewables such as wind and solar have shown steady increase. ERCOT's nuclear generation and hydroelectric power have both remained stable, though we should note that hydroelectric has very minimal contribution.

```{r Plot TS Fuels}
ts_data <- data.frame(
  time = time(ts_load),
  load = ts_load,
  temperature = ts_temp,
  coal_and_lignite = ts_coal_and_lignite,
  hydro = ts_hydro,
  nuclear = ts_nuclear,
  solar = ts_solar,
  wind = ts_wind,
  natural_gas = ts_natural_gas
)

autoplot(ts_natural_gas, series = "Natural Gas", size = 0.5, alpha = 0.6) +
  autolayer(ts_coal_and_lignite, series = "Coal and Lignite", size = 0.5, alpha = 0.8) +
  autolayer(ts_wind, series = "Wind", size = 0.3, alpha = 0.8) +
  autolayer(ts_nuclear, series = "Nuclear", size = 0.5, alpha = 0.8) +
  autolayer(ts_solar, series = "Solar", size = 0.5, alpha = 0.6) +
  autolayer(ts_hydro, series = "Hydro", size = 0.5, alpha = 0.6) +
  ggtitle("ERCOT Fuel Mix") +
  xlab("Time") +
  ylab("Load (MW)") +
  scale_color_manual(values = c("red", "darkgreen", "skyblue2", "midnightblue", "orange", "pink")) + 
  theme_minimal() +
  guides(colour = guide_legend(title = "Time Series"))
```
While several fuels make up the ERCOT fuel mix, natural gas seems the most relevant for our acf/pacf analysis, as its time series plot mirrors the shape of the ERCOT load. The ACF and PACF of natural gas prices also closely resemble those of ERCOT load and temperature, showing a gradual decay in the ACF and irregular spikes in the PACF. This suggests that natural gas load exhibits non-stationary behavior with seasonal components, highlighting its impact in shaping electricity demand patterns and making it an important variable in scenario generation and forecasting.

# ACF and PACF for Natural Gas

```{r ACF and PACF for Natural Gas}
# ACF and PACF for Natural Gas
ggAcf(ts_natural_gas, lag.max = 365) + ggtitle("ACF of Natural Gas") + theme_minimal()
ggPacf(ts_natural_gas, lag.max = 365) + ggtitle("PACF of Natural Gas") + theme_minimal()
```

# Methodology and Models

Modeling ERCOT load is inherently complex due to the presence of multiple seasonal patterns, non-linearities, and variability driven by weather, human behavior, and market conditions. To capture these dynamics, we used a varity of time series models, each offering unique strengths.