---
title: "ERCOT Load Forecasting"
subtitle: https://github.com/alex-lopez70/ChuangWhiteheadLopez_ENV797_TSA_FinalProject/
author: "Daniel Whitehead, Alex Lopez, Jessalyn Chuang"
output: pdf_document
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#tinytex::install_tinytex()
```

```{r, include=FALSE}
library(rvest);library(dataRetrieval);library(tidyverse);library(lubridate);library(viridis)
library(here);library(httr);library(jsonlite);library(knitr);library(kableExtra)
library(ggplot2);library(forecast);library(Kendall);library(tseries);library(outliers)
library(tidyverse);library(smooth);library(zoo)
here()
```

```{r data pull, include=FALSE}
#read csv files
ERCOT_fuelmix_2017to2018 <- read.csv(file="./Raw_Data/ERCOT_fuel_mix_112017_12312018.csv")
ERCOT_fuelmix_2019to2021 <- read.csv("./Raw_Data/ERCOT_fuel_mix_112019_12312021.csv")
ERCOT_fuelmix_2022to2025 <- read.csv("./Raw_Data/ERCOT_fuel_mix_112022_442025.csv")
ERCOT_fuelmix <- rbind(
  ERCOT_fuelmix_2017to2018,
  ERCOT_fuelmix_2019to2021,
  ERCOT_fuelmix_2022to2025
)


ERCOT_load_2017to2019 <- read.csv("./Raw_Data/ERCOT_load_112017_12312019.csv")
ERCOT_load_2020to2022 <- read.csv("./Raw_Data/ERCOT_load_112020_12312022.csv")
ERCOT_load_2023to2025 <- read.csv("./Raw_Data/ERCOT_load_112023_442025.csv")
ERCOT_load <- rbind(
  ERCOT_load_2017to2019,
  ERCOT_load_2020to2022,
  ERCOT_load_2023to2025
)

ERCOT_Temp_2017to2019 <- read.csv("./Raw_Data/ERCOT_Temp_112017_112020.csv")
ERCOT_Temp_2020to2023 <- read.csv("./Raw_Data/ERCOT_Temp_112020_112023.csv")
ERCOT_Temp_2023to2025 <- read.csv("./Raw_Data/ERCOT_Temp_112023_452025.csv")
ERCOT_temp <- rbind(
  ERCOT_Temp_2017to2019,
  ERCOT_Temp_2020to2023,
  ERCOT_Temp_2023to2025
)

```

```{r ts object creation, include=FALSE}
ERCOT_fuelmix$Date <- as.Date(ERCOT_fuelmix$interval_start_local)
ERCOT_load$Date <- as.Date(ERCOT_load$interval_start_local)

ERCOT_temp$Date <- as.Date(ERCOT_temp$interval_start_local)
ERCOT_temp$Average <- rowMeans(ERCOT_temp[,7:14], na.rm = FALSE, dims = 1)
ERCOT_temp_daily <- ERCOT_temp %>% 
  group_by(Date) %>% 
  summarise(mean(Average)) %>% 
  ungroup()

ts_load <- msts(ERCOT_load$load, seasonal.periods = c(7,365.25), 
                start = c(2017,1,1))
ts_temp <- msts(ERCOT_temp_daily[,2], seasonal.periods = c(7,365.25), 
                start = c(2017,1,1))
ts_coal_and_lignite <- msts(ERCOT_fuelmix$coal_and_lignite, seasonal.periods = c(7,365.25), 
                start = c(2017,1,1))
ts_hydro <- msts(ERCOT_fuelmix$hydro, seasonal.periods = c(7,365.25), 
                start = c(2017,1,1))
ts_nuclear <- msts(ERCOT_fuelmix$nuclear, seasonal.periods = c(7,365.25), 
                start = c(2017,1,1))
ts_solar <- msts(ERCOT_fuelmix$solar, seasonal.periods = c(7,365.25), 
                start = c(2017,1,1))
ts_wind <- msts(ERCOT_fuelmix$wind, seasonal.periods = c(7,365.25), 
                start = c(2017,1,1))
ts_natural_gas <- msts(ERCOT_fuelmix$natural_gas, seasonal.periods = c(7,365.25), 
                start = c(2017,1,1))

```

```{r train/test split, include=FALSE}
n_for <-  365

ts_load_train <- subset(ts_load,end = length(ts_load)-n_for) #stops 365 days before

#create a subset for testing purpose
ts_load_test <- subset(ts_load, start = length(ts_load)-n_for) #last 365 observations
```

# Introduction and Motivation
Load growth used to be fairly stagnant and was managed by implementing more efficiency measures (Walton, 2023). However, with innovation in electrical technologies (EVs, heat pumps, etc.) and the introduction of new load sources (crypto mining and data centers), electrical load has skyrocketed in recent years. As a result, we need new and innovative measures of responding to this load growth, such as demand response, interconnection reform, among others. The ability to forecast load growth is an interesting task, because there are both general trends (growth in electric technologies, climate change, and increasing electrification, among others) and seasonal trends (periodic variations in energy demand and consumption) that impact load.

# Relevance and Objectives
In the US, ERCOT is a market witnessing tremendous load growth. In 2018, summer peak load record was 69.5 GW, but in just 5 years, this record was broken with a 16 GW increase, with the new record hitting 85.6 GW (Wilson and Zimmerman, 2023). Moreover, ERCOT set a new winter demand peak in February 2025 with load being above 80 GW for the first time, similar figures usually seen in summer months, which highlights load growth not being exclusive to the summer (Kleckner, 2025). A lot of this can be attributed to crypto mining facilities, data centers, and industrial electrification (Guo, 2025).

At the same time, ERCOT has had some really exciting developments, from being the wholesale market with the greatest growth in solar PV and battery storage, to having the fastest interconnection process out of all the ISO/RTOs in the US, making ERCOT one of the most dynamic electric wholesale markets in North America. These developments have all helped ERCOT respond to load growth in the wake of extreme weather events like Winter Storm Uri in 2021. In this context, being able to better forecast load growth in ERCOT will help contribute to greater innovation in policy and technology to support load growth while continuing to reliably serve pre-existing load.

# Dataset Information
The ERCOT load dataset was collected from GridStatus.io, a website and Python library that provides a uniform API for accessing electricity supply, demand, and pricing data for the major ISOs in the United States (https://www.gridstatus.io/live). 

## Description
Using its “Export Data” tool, we queried hourly temperature and daily load and fuel mix data for ERCOT from January 2017 to April 2025. Below is a description of the data set retrieved: 

```{r echo=FALSE}
#data set description
dataset_description <- data.frame(
  Column1 = c("Units", "Frequency", "Subsets"),
  Load = c("MW", "Daily", ""),
  Temperature = c("Fahrenheit", "Hourly", "ERCOT Zone"),
  FuelMix = c("MW", "Daily", "Fuel (Coal and lignite, Hydro, Nuclear,
                 Solar, Wind, Natural Gas)")
)

dataset_description$Column1 <- cell_spec(dataset_description$Column1, bold = TRUE)

# Create the kable table with the first column bolded
kbl(dataset_description, caption = "Data Set Description", 
    col.names = c("", "Load", "Temperature", "Fuel Mix"), escape = FALSE) %>%
  kable_styling(full_width = FALSE, position = "center", latex_options = "hold_position") %>%
  kable_styling(latex_options="striped")

```

## Data Wrangling
The earliest available data for ERCOT fuel mix was from January 2017, so we decided to retrieve load, temperature, and fuel mix data from January 2017 onwards. After importing each of these three data sets (daily values except for temperature data) for the dates 1 January 2017 to 4 April 2025 and converting them to data frames, we created a new 'Date' column for each of the three date sets that converted the time given in the 'interval_start_local' to a YYYY-MM-DD format. Regarding the temperature data, since it is hourly data divided by zone, we first found the hourly average for each zone, then the daily average for each zone, and finally we found the daily average using all the zones in ERCOT by using rowMeans(), group_by(), and summarise().

Since we are working with higher frequency data with multiple seasonalities, we are using the `msts` class from package `forecast` to create time series objects for our daily data sets. Unlike the ts() function, reserved for data exhibiting single seasonality, the msts() function allows for non-integer frequency specification. We therefore created eight time series objects, one for load, one for temperature, and one for each of the fuels we are analyzing, namely coal and lignite, hydro, nuclear, solar, wind, and natural gas, with a 'seasonal.period = c(7, 365.25)' for each. Below are summaries of the data structures of each of the three wrangled data frames (for load, temperature, and fuel mix) that were then used to create time series objects by using msts():

```{r data.summary, echo=FALSE}
ERCOT_load_summary <- data.frame(
  Variable = c('interval_start_local', 'interval_start_utc', 'interval_end_local',
               'interval_end_utc', 'load', 'Date'),
  Description = c('Start Time for the Day', 'Start Time for the Day (UTC)',
                  'End Time for the Day', 'End Time for the Day (UTC)', 'Load (MW)',
                  'Date (YYYY-MM-DD)'),
  Units = c('Character', 'Character', 'Character', 'Character', 'Numeric', 
            'Date')
)

ERCOT_temp_summary <- data.frame(
  Variable = c('Date', 'mean(Average)'),
  Description = c('Date (YYYY-MM-DD)', 'Average Daily Temperature (Fahrenheit)'),
  Units = c('Date', 'Numeric')
)

ERCOT_fuelmix_summary <- data.frame(
  Variable = c('interval_start_local', 'interval_start_utc', 'interval_end_local',
               'interval_end_utc', 'coal_and_lignite', 'hydro', 'nuclear', 
               'solar', 'wind', 'natural_gas'),
  Description = c('Start Time for the Day', 'Start Time for the Day (UTC)',
                  'End Time for the Day', 'End Time for the Day (UTC)',
                  'Generation from Coal and Lignite (MW)',
                  'Generation from Hydro (MW)', 'Generation from Nuclear (MW)',
                  'Generation from Solar (MW)', 'Generation from Wind (MW)',
                  'Generation from Natural Gas (MW)'),
  Units = c('Character', 'Character', 'Character', 'Character', 'Numeric', 
            'Numeric', 'Numeric', 'Numeric', 'Numeric', 'Numeric')
)

```

```{r table1, echo=FALSE}
kable(ERCOT_load_summary, caption = "ERCOT Load Summary")
```


```{r table2, echo=FALSE}
kable(ERCOT_temp_summary, caption = "ERCOT Daily Temperature Summary")
```


```{r table3, echo=FALSE}
kable(ERCOT_fuelmix_summary, caption = "ERCOT Fuel Mix Summary")
```

# Data Exploration

## ERCOT Load

## ERCOT Temperature

## ERCOT Fuel Mix

# Methodology and Models

## STL+ETS

## TBATS

## ARIMA + Fourier Terms

## Neural Network

## SARIMA

## SSES

## SSBSM

## Scenario Generation

# Results

# Conclusions

# References
1. Guo, Kayla. Data centers are booming in Texas. What does that mean for the grid? Published 24 Jan. 2025. https://www.texastribune.org/2025/01/24/texas-data-center-boom-grid/.

2. Kleckner, Tom. Batteries, Solar Help ERCOT Meet Record Winter Peak, 20 Feb. 2025, https://www.rtoinsider.com/98816-batteries-solar-help-ercot-meet-record-winter-peak/. 

3. Walton, Robert. “US Electricity Load Growth Forecast Jumps 81% Led by Data Centers, Industry: Grid Strategies.” Utility Dive, 13 Dec. 2023, www.utilitydive.com/news/electricity-load-growing-twice-as-fast-as-expected-Grid-Strategies-report/702366/.

4. Zimmerman, Zach, and John  D Wilson. The Era of Flat Power Demand Is Over, Dec. 2023, gridstrategiesllc.com/wp-content/uploads/2023/12/National-Load-Growth-Report-2023.pdf. 